% Section .............................................................................................................
\section{Two Layers in Every Physical Model}
\label{section:layers}

The collapse calculation exemplifies a general structure present in all physical models, but rarely made explicit. Every model carries at least two distinguishable layers of content:

\begin{enumerate}
  \item \textbf{A predictive-mathematical layer:} the equations, boundary conditions, and inference rules that generate quantitative predictions about measurable quantities.
  \item \textbf{An ontological layer:} the implicit commitments about what kinds of entities exist, what properties they possess independently of measurement, and how they relate causally.
\end{enumerate}

The atom is our primary case study, but the diagnostic generalises:
we identify the same two-layer structure in the ultraviolet
catastrophe and in the quantum measurement problem.

Standard empiricist philosophy of science focuses almost exclusively on the first layer \cite{vanfraassen_1980_image}. A model is adequate if its predictions agree with experiment; inadequate otherwise. On this view, the classical atomic model fails because it predicts a collapse timescale of $\sim10^{-11}$\,s while atoms are stable indefinitely. That diagnosis is correct, but incomplete.

% Subsection ..........................................................................................................
\subsection{The Ontological Layer of the Classical Model}
\label{subsection:layers}

The deeper diagnosis is that the classical model fails at the second layer. Newton's equations for dynamics, Coulomb's law for electromagnetism, and Larmor's radiation formula are not wrong in themselves---they remain excellent approximations for macroscopic charged objects. Yet, they fail for the electron in the hydrogen atom because they embed a specific ontology:

\begin{itemize}
  \item Electrons are \emph{point particles} with definite position and
    momentum at every instant.
  \item They move along \emph{continuous trajectories} governed by
    deterministic force laws.
  \item They act as \emph{classical sources} for the electromagnetic
    field, coupling through their charge and acceleration.
\end{itemize}

It is known nowadays that each of these assumptions is false. A model can therefore be \emph{mathematically consistent}, \emph{empirically adequate in its domain}, and \emph{ontologically erroneous}. The classical hydrogen atom is a case study in exactly this combination.



% Subsection ..........................................................................................................
\subsection{The General Diagnostic Distinction}
\label{subsection:distinction}

We propose the following general principle:

\begin{quote}
  \emph{When a physical model fails empirically at the boundary of its domain, one must ask whether the failure is predictive or ontological. A predictive failure can often be corrected by modifying equations within the same ontological framework. An ontological failure requires replacing the picture of what exists.}
\end{quote}

As an example, the transition from Newtonian to relativistic mechanics is largely \emph{predictive}: the ontology of point particles with definite trajectories is preserved; only force laws and the metric of spacetime change. The transition from classical to quantum mechanics is \emph{ontological}: the very category of ``particle with a definite trajectory'' ceases to apply.

\medskip

This distinction has implications for how we interpret theoretical progress. Kuhnian paradigm shifts \cite{kuhn_1962_structure} and Lakatosian research programmes \cite{lakatos_1978_methodology} both capture aspects of it, but neither systematically separates the predictive from the ontological layer. The collapse calculation makes the separation vivid: the predictive machinery is flawless; it is the ontological presuppositions that must go.

% Al final de la sección 2.2, tras el párrafo sobre Newton→Einstein→QM


\subsection{A Second Case: The Ultraviolet Catastrophe}

The atomic collapse is not an isolated case.
The ultraviolet catastrophe \cite{rayleigh_1900_radiation,
jeans_1905_partition} presents the same diagnostic structure.
Applying the equipartition theorem to the electromagnetic field modes
in a cavity yields a spectral energy density $u(\nu,T) \propto \nu^2
k_BT$, which diverges as $\nu \to\infty$---a prediction that is
mathematically exact given classical statistical mechanics and
electrodynamics.
Wien's empirical law \cite{wien_1896_radiation} and ultimately
Planck's quantum hypothesis \cite{planck_1900_radiation} resolved the
discrepancy not by correcting an equation within the classical
framework but by replacing its ontological foundation: energy
exchange between matter and field is not continuous but quantised in
units $E = h\nu$.
The failure, again, was not predictive within a domain but
ontological at the domain boundary.

\medskip

\TODO{added, pero afinar: Expand: Cartwright on laws vs.\ capacities; Giere on model-world relations; Chakravartty on semi-realism. Key point: none of these frameworks isolates the ontological layer as a separate diagnostic target in quite the way we propose.}

% -----------------------------------------------------------
% INSERT after the paragraph ending "...ontological presuppositions that must go."
% Replaces the \TODO block
% -----------------------------------------------------------

The distinction we draw is related to, but distinct from, three influential
positions in the philosophy of science that touch on the limits of physical
models.

Cartwright's argument in \emph{How the Laws of Physics Lie}
\cite{cartwright_1983_laws} is the closest precedent: she contends that
the fundamental laws of physics achieve their generality by systematically
misrepresenting causal capacities, and that models introduce
\emph{ad hoc} corrections that restore empirical accuracy at the cost of
ontological transparency.
Our claim is more specific: the classical atomic model does not merely
misrepresent capacities---it presupposes a \emph{category} of entity
(the classically trajectoried particle) that simply does not exist at
the relevant scale.
The failure is not one of approximation but of kind.

Giere's model-theoretic account \cite{giere_1988_science} holds that
models are abstract objects that \emph{represent} target systems through
a relation of similarity, with no claim to literal truth about underlying
ontology.
On this view, asking whether the classical atomic model ``has the right
ontology'' is a category error: models represent, they do not describe.
We resist this conclusion.
The collapse calculation shows that the classical model does not merely
\emph{fail to represent} the hydrogen atom; it \emph{entails} a
prediction---sixteen picoseconds to collapse---that is straightforwardly
and catastrophically false.
That is not a similarity failure; it is an ontological commitment that
nature rejects.

Chakravartty's semi-realism \cite{chakravartty_2007_metaphysics}
distinguishes \emph{detection properties}---those causally responsible
for our measurements---from \emph{auxiliary properties} that models
posit but that may not correspond to anything real.
This is the closest structural analogue to our two-layer distinction.
The difference is diagnostic emphasis: Chakravartty's framework is
designed to identify what to be realist \emph{about} within a successful
theory, whereas ours is designed to identify \emph{why} a theory fails
at a domain boundary.
The classical collapse is a case where the auxiliary properties
(definite position, continuous trajectory, classical field coupling)
are not merely unconfirmed but actively contradicted.
None of these three frameworks isolates the ontological layer as a
separate \emph{diagnostic target} at the moment of theoretical failure
in quite the way we propose.